Multisequence algorithm for coarse-grained biomolecular simulations: application to protein fold switching

### 2017-05-23 Manuscript.tex

\begin{abstract}
We evaluate a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that allows the thermodynamics of multiple sequences to be calculated in a single ``multisequence" run. By carrying out a random walk in sequence space, conformational sampling is also enhanced. Escape from local energy minima is achieved by visiting sequences for which the minima is shallower or absent. We apply the algorithm alongside a coarse-grained model for protein folding with 7 atoms per amino acid to sets of sequences with up to 35 amino acids spanning a large sequence space. With the multisequence algorithm, we show that the folding thermodynamics of $>$ 1,000 sequences at relatively low temperatures can be readily calculated. We use the resulting data to explore the statistics of pathways in sequence space between two pairs of folds.

@@

\subsection{Stability measures}
\noindent
To quantify the stabilities of various folds, we use the measures introduced in Ref.~\cite{Holzgrafe2015}. We classify conformations $r$ as part of a fold if $Q\ge0.8$ where $Q\in [0,1]$ quantifies the similarity of $r$ to that fold.

For folds A and B, $Q$ is determined by defining a set of backbone-backbone hydrogen bonds that are present in those folds (ignoring N and C terminal amino acids). For fold A, the fold-specific set of contacts (i,j) between the CO group of amino acid i and the NH group of amino acid j is $\{$(2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15)$\}$, and for fold B the contact set is $\{$(3,14), (5,12), (7,10), (10,7), (12,5), (14,3)$\}$. A hydrogen bond is considered formed if (i) the NHO and COH angles are both $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

%\begin{equation}
%Q_\mathrm{2\alpha} = ( Q_\alpha^\mathrm{1-16} + Q_\alpha^\mathrm{20-35} + Q_\mathrm{tert}) ) / 3
%\end{equation}
%
%\begin{equation}
%Q_\mathrm{\alpha\beta} = ( Q_\alpha^\mathrm{1-16} + Q_\beta^\mathrm{20-35} + Q_\mathrm{tert} ) / 3
%\end{equation}

@@

\vspace{12pt}
\subsection{A1-N1 and A2-TN binary sequence space}
\noindent
We now apply the MS algorithm to the sets {\SI} and {\SII} which represent the full binary sequence space for the A1-N1 and A2-TN systems, respectively. Each set includes $2^{10}=1,024$ sequences. Using MS simulations, we determined the thermodynamic behavior at the temperatures $\TlowI$ and $\TlowII$ for the A1-N1 and A2-TN systems, respectively (see Table~1). In particular, this allows us to find the stabilities relative to folds IA and IB ($\PaI$ and $\PbI$) for all sequences in {\SI} and the stabilities relative to folds IIA and IIB ($\PaII$ and $\PbII$) for all sequences in {\SII}.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to IA-IB or IIA-IIB fold changes without passing through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at only one amino acid position. To determine if a sequence is stable we used the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PaI + \PbI$ and $\PaII + \PbII$ for A1-N1 and A2-TN, respectively. The precise network obtained depends, of course, on the cut-off value of $\Pcut$. In general, a higher $\Pcut$ means a selection of more stable pathways. Fig.~5 shows the networks obtained with $\Pcut=0.50$. The number of viable mutational pathways that connect the start and end points are 516,972 between A1 and N1 and  57,912 between A2 and TN. Because there are $10! =3,628,800$ possible mutational pathways between start and end points in both cases, these pathways represent 14.2 \% and 1.6 \% of all possible pathways in A1-N1 and A2-TN, respectively. Hence, A1 and N1 are rather highly connected at this stability threshold. For $\Pcut=0.60$, the numbers are 104,640 pathways (2.9\%) for A1-N1 and 22,512 (0.6\%) pathways for A2-TN. We find that there are no possible A1-to-N1 or A2-to-TN pathways when $\Pcut\ge0.74$ and $\Pcut\ge 0.66$, respectively. \\

@@

\section{Discussion}
\noindent
We have tested an algorithm for coarse-grained biomolecular simulations that relies on making the sequence of the molecule a dynamical parameter.

For comparison, we used the well-established generalized-ensemble ST algorithm and focused on 144 model sequences with 16 amino acids ({\SR}). As a measure of computational efficiency, we used the statistical error $\sigE$ in the estimation of the average total energy $E$ for 4 different sequences which are part of {\SR}.

Our results show that at high $T$, well above the folding temperatures of both A1 and N1, we find that the statistical errors $\sigE$ for ST and MS are small and very similar. This is not surprising given that conformational sampling of relatively short polymer chains at high $T$ does not involve crossing major energy barriers. Therefore, successive sampled conformational states, for a given combination of sequence $\sbar$ and temperature $T$, are likely uncorrelated in both ST and MS simulations which, in turn, leads to similar estimates of $\Eave$ (and $\sigE$). At low $T$,...

Hence, we conclude that the computational efficiency of the MS algorithm is comparable to ST. However, that while the ST simulation requires that the thermodynamic behavior is calculated at a range of temperatures, MS works well on its own directly at low temperatures. ST was introduced as an extension to the standard, fixed-$T$ Metropolis-Hastings algorithm in order to overcome the poor mixing of states at low temperatures, i.e., the tendency for the simulations at low $T$ to become trapped in local free energy minima. The MS algorithm, however, efficiently calculates the thermodynamics at low $T$'s without visits to other $T$'s. In the MS algorithm, it is visits to sequences with poor native state stability rather than high $T$'s that drive the decorrelation of generated states and escape from local traps. This way, calculation of low-$T$ thermodynamics is enhanced in MS simulations while at the same time enhancing coverage in sequence space. This makes the MS algorithm ideal for situations in which the finite-$T$ thermodynamic behavior of multiple sequences are of interest.

A notable feature of the MS algorithm is that it provides escape from local energy minima on the free energy landscape of proteins without resorting to a change in temperature. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in lipid membranes, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Sampling in membrane simulations at high $T$, such as in replica-exchange simulations, are usually avoided because it can lead to unwanted perturbation of the lipid bilayer structure although it can be a viable approach in particular cases.~\cite{Ulmschneider2010}

>>>>

\begin{abstract}

We consider a generalized-ensemble algorithm for coarse-grained simulations of biomolecules, which allows the thermodynamic behaviors of two or more sequences to be determined in a single multisequence run. By carrying out a random walk in sequence space, the method also enhances conformational sampling. Escape from local energy minima is accelerated by visiting sequences for which the minima are more shallow or absent. We apply the method alongside an intermediate-resolution coarse-grained model for protein folding with 7 atoms per amino acids. The potential for large-scale coverage of sequence space is explored by applying the method to two sets with $>$1,000 sequences each. The resulting thermodynamic data is used to explore the sequence-structure and sequence-stability relationships between pairs of protein folds with different secondary structures. The potential implications for the evolution of new protein folds are discussed.

@@

\subsection{Observables}
\noindent
Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and also described briefly below. We define first two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating what fraction of the fold-specific backbone-backbone hydrogen bonds have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The  stabilities of fold IA and IB are then defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.80)$, respectively, i.e., stability is the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.xx$ for A1 and $\PIB=0.xx$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert}) ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $C_\beta$-$C_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} As for IA and IB, the stabilities of folds IIA and IIB are defined as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.80)$, respectively. 
%In calculating $Q$ values, a hydrogen bond is considered formed if (i) the NHO and COH angles are both $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

@@

\vspace{12pt}
\subsection{A1-N1 and A2-TN binary sequence space}
\noindent

We now apply the MS algorithm to the sets {\SI} and {\SII} which represent the full binary sequence space for the A1-N1 and A2-TN systems, respectively. Each set includes $2^{10}=1,024$ sequences. Using MS simulations, we determined the thermodynamic behavior at the temperatures $\TlowI$ and $\TlowII$ for the A1-N1 and A2-TN systems, respectively (see Table~1). In particular, this allows us to find the stabilities relative to folds IA and IB ($\PIA$ and $\PIB$) for all sequences in {\SI} and the stabilities relative to folds IIA and IIB ($\PIIA$ and $\PIIB$) for all sequences in {\SII}.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to IA-IB or IIA-IIB fold changes without passing through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at only one amino acid position. To determine if a sequence is stable we used the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PIA + \PIB$ and $\PIIA + \PIIB$ for A1-N1 and A2-TN, respectively. The precise network obtained depends, of course, on the cut-off value of $\Pcut$. In general, a higher $\Pcut$ means a selection of more stable pathways. Fig.~5 shows the networks obtained with $\Pcut=0.50$. The number of viable mutational pathways that connect the start and end points are 516,972 between A1 and N1 and  57,912 between A2 and TN. Because there are $10! =3,628,800$ possible mutational pathways between start and end points in both cases, these pathways represent 14.2 \% and 1.6 \% of all possible pathways in A1-N1 and A2-TN, respectively. Hence, A1 and N1 are rather highly connected at this stability threshold. For $\Pcut=0.60$, the numbers are 104,640 pathways (2.9\%) for A1-N1 and 22,512 (0.6\%) pathways for A2-TN. We find that there are no possible A1-to-N1 or A2-to-TN pathways when $\Pcut\ge0.74$ and $\Pcut\ge 0.66$, respectively. \\

@@

\section{Discussion}
\noindent
We have tested a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that works by making the sequence a dynamic parameter, thereby leading to a random walk in sequence space. This approach was evaluated for protein folding simulations by comparing with ST. The results indicate two main benefits of the MS approach. On the one hand, it provides a convenient way to sample the respective canonical distributions for a large set of sequences in a single run and, on the other, it also enhances the sampling of conformational space.

The conformational sampling efficiency of the MS algorithm can be assessed from the comparison with ST in Figure~3. Although there is no single ``fair" way to compare MS and ST, we chose as a measure of efficiency the estimated statistical error on the total energy, $\sigE$, obtained with roughly the same computational cost per temperature and sequence. At the highest studied temperatures, which are well above the folding temperatures of both A1 and N1, the statistical errors $\sigE$ are basically the same for the two methods. This is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations for a given combination of sequence $\sbar$ and temperature $T$, are therefore likely mostly uncorrelated in both methods leading to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling vis-{\'a}-vis ST is achieved despite that simulations are carried out at constant temperature. Hence, rather than promoting escape from local minima in the free energy landscape by visits to higher temperature, as is the case in  ST~\cite{Marinari1992,Lyubartsev1992} or replica-exchange,~\cite{Swendsen1986}  MS simulations escape local minima through visits to sequences for which the minima may not be as deep or are altogether absent. For the same reason, the performance of the MS algorithm will depend on the size and character of sequences included. In particular, the MS simulation likely hinge on the inclusion of at least some sequences with poor stabilities such that at least partial unfolding of the chain is regularly triggered and promoting transitions to new conformational states.

The MS method should not be seen as a general method to speed up conformational sampling due to the complexity of performing sequence updates in most biomolecular models. However, our results indicate that for coarse-grained models that permit the sequence updates as a Metropolis step and when visits to higher $T$ is unwanted (or unnecessary), the MS method can be highly efficient. A notable feature of the algorithm is that it provides escape from local minima on the free energy landscape of proteins without resorting to a change in $T$. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in the context or ordered dense phases such as lipid bilayers, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Simulations of lipid bilayers at elevated $T$ must typically be avoided because it can lead to unwanted perturbations of the membrane structure, although it can be a viable approach to speed up conformational sampling in particular cases.~\cite{Ulmschneider2010}

%Overall our results show that, when technically feasible, a multisequence approach can be highly efficient for calculating the thermodynamics of a large number of amino acid sequences even at low temperatures. To demonstrate that the method to large sequence sets we applied it to the full binary sequences spaces between the model sequence pairs A1-N1 and A2-TN, which both include 1,024 sequences. These calculations would have been difficult to carry out ``in series" using some other standard method such as simulated tempering.

%For comparison, we used the well-established generalized-ensemble ST algorithm and focused on 144 model sequences with 16 amino acids ({\SR}). As a measure of computational efficiency, we used the statistical error $\sigE$ in the estimation of the average total energy $E$ for 4 different sequences which are part of {\SR}.

%Hence, we conclude that the computational efficiency of the MS algorithm is comparable to ST. However, that while the ST simulation requires that the thermodynamic behavior is calculated at a range of temperatures, MS works well on its own directly at low temperatures. ST was introduced as an extension to the standard, fixed-$T$ Metropolis-Hastings algorithm in order to overcome the poor mixing of states at low temperatures, i.e., the tendency for the simulations at low $T$ to become trapped in local free energy minima. The MS algorithm, however, efficiently calculates the thermodynamics at low $T$'s without visits to other $T$'s. In the MS algorithm, it is visits to sequences with poor native state stability rather than high $T$'s that drive the decorrelation of generated states and escape from local traps. This way, calculation of low-$T$ thermodynamics is enhanced in MS simulations while at the same time enhancing coverage in sequence space. This makes the MS algorithm ideal for situations in which the finite-$T$ thermodynamic behavior of multiple sequences are of interest.

%A notable feature of the MS algorithm is that provides escape from local energy minima on the free energy landscape of proteins without resorting to a change in temperature. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in lipid membranes, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Sampling in membrane simulations at high $T$, such as in replica-exchange simulations, are usually avoided in membrane simulations because it can lead to unwanted perturbation of the lipid bilayer structure although it can be a viable approach in particular cases.~\cite{Ulmschneider2010}

%A notable feature of the MS algorithm is that it provides escape from local energy minima on the free energy landscape of proteins without resorting to a change in temperature. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in lipid membranes, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Sampling in membrane simulations at high $T$, such as in replica-exchange simulations, are usually avoided because it can lead to unwanted perturbation of the lipid bilayer structure although it can be a viable approach in particular cases.~\cite{Ulmschneider2010}
